Prediction of PIK3CA mutation with gene expression

### 2020-04-23 plosOne/plosOne.Rmd
rpackage.bib

---
title: Prediction of PIK3CA mutation with gene expression
author:
  - name: Jun Kang
    email: jkang.alien@gmail.com
    affiliation: Department of Hospital Pathology, Seoul St. Mary’s Hospital, College of Medicine, The Catholic University of Korea, Seoul, South Korea
    corresponding: jkang.alien@gmail.com
abstract: |
bibliography: [mybibfile.bib, pik3ca.bib] 
indent: true
output: rticles::plos_article
csl: plos.csl
---

@@

We used TCGA pancancer dataset. PIK3CA mutation data was get using cgdsr rpackage.[@jacobsenCgdsrRbasedAPI2019] Gene expression data was get from GDAC firehose using RTCGAToolbox R package. [@samurRTCGAToolboxNewTool2014] ER immunostain postive HER2 immunostain negative and/or SISH negative breast cancers are included. Data of invasive ductal carcinomas were used for training set and data of invasive lobular carcinoma were used for test set. Number of observations were 530 in training set and 188 in test set.

@@

To narrow down potential predictors, Genes with a large the median absolute deviation (more than third-quartiles) were selected. 5000 out of 20502 genes were included in the modeling process. Yeo-Johnson transformation was done to correct skewness. Centering and scaling were done. All preprocessing was done using recipe r package. [@kuhnRecipesPreprocessingTools2020] Penalized logistic regression was applied to prediction modeling. 10-fold cross-validation with targe variable stratification was done over the hyperparameter grid: $\lambda$ {$10^{-5}$, $10^{-4}$,$10^{-3}$,$10^{-2}$,$10^{-1}$, $10^{0}$}, $\alpha$ {0.0, 0.25, 0.5, 0.75}. Model performance was evealuated with the area under receiver operating characteristic (ROC) and area under the ROC (AUROC) and precision recall (PR) and area under curve (AUPR). The modelin process was done with tidymodels rpackage.  [@kuhnTidymodelsEasilyInstall2020]

@@

## Selecting model

Cross-validation  
The model showed best performance at lambda = 0.01 and alpha = 1.0 (Ridge regression).

## Prediction performance

@@

## Performance of each cancer type  
Because tht prevalence of PIK3CA mutation is vary across the cancer type, the performance of each cancer type was investigated. The AUROC and AUPR were positively correlated between train set and test set in cancer type subanalysis. (Figure 1B) The AUPR was high in cancer type with high PIK3CA mutation rate such as colon, brest, Uterus cancer types. The AUROC did not correlated with PIK3CA mutation rate of each cancer types. (Figure 1C)

@@

@@

>>>>

---
title: Prediction of PIK3CA mutation with gene expression
author:
  - name: Jun Kang
    email: jkang.alien@gmail.com
    affiliation: Department of Hospital Pathology, Seoul St. Mary’s Hospital, College of Medicine, The Catholic University of Korea, Seoul, South Korea
    corresponding: jkang.alien@gmail.com
abstract: |
bibliography: [mybibfile.bib, pik3ca.bib, rpackage.bib] 
indent: true
output: rticles::plos_article
csl: plos.csl
---

@@

We used TCGA pancancer dataset. PIK3CA mutation data was get using cgdsr rpackage.[@jacobsenCgdsrRbasedAPI2019] Gene expression data was get from GDAC firehose using RTCGAToolbox R package. [@samurRTCGAToolboxNewTool2014]

@@

To narrow down potential predictors, Genes with a large the median absolute deviation (more than third-quartiles) were selected. 5000 out of 20502 genes were included in the modeling process. Yeo-Johnson transformation was done to correct skewness. Centering and scaling were done. All preprocessing was done using recipe r package. [@kuhnRecipesPreprocessingTools2020] Penalized logistic regression was applied to prediction modeling. 10-fold cross-validation with targe variable stratification was done over the hyperparameter grid: $\lambda$ {$10^{-5}$, $10^{-4}$,$10^{-3}$,$10^{-2}$,$10^{-1}$, $10^{0}$}, $\alpha$ {0.0, 0.25, 0.5, 0.75}. Lambda is  penalty scaling parameter and alpha is mixing parameter of penalty function ($(1-\alpha)/2 \lVert\beta\rVert_2^2+\alpha\lVert \beta \rVert_1$). [@friedmanRegularizationPathsGeneralized2010] Model performance was evealuated with the area under receiver operating characteristic (ROC) and area under the ROC (AUROC) and precision recall (PR) and area under curve (AUPR). The modelin process was done with tidymodels rpackage.  [@kuhnTidymodelsEasilyInstall2020]

@@

## Selecting model and performance estimation

In 10-fold cross-validation, the model with lambda = 0.01 and alpha = 1.0 (Ridge regression) were best performance in terms of AUROC. The final model was trained with the selected hyperparameters with all trainset.

@@

## Performance of each cancer type

Because tht prevalence of PIK3CA mutation is vary across the cancer type, the performance of each cancer type was investigated. The AUROC and AUPR were positively correlated between train set and test set in cancer type subanalysis. (Figure 1B) The AUPR was high in cancer type with high PIK3CA mutation rate such as colon, brest, Uterus cancer types. The AUROC did not correlated with PIK3CA mutation rate of each cancer types. (Figure 1C)

@@

OUr model includes cancer type predictor and they are stronger than gene expression data. The varing prevalence of PIC3CA mutation across cancer type might reason of the strong cancer type predictor. If the cancer type was wrong or can not be determined, our model performance can be poor.

@@

- Proof of concept study  
- Reproducibility of gene expression prediction model

### 2020-04-23 plosOne/plosOne.tex
rpackage.bib

We used TCGA pancancer dataset. PIK3CA mutation data was get using cgdsr
rpackage.{[}3{]} Gene expression data was get from GDAC firehose using
RTCGAToolbox R package. {[}4{]} ER immunostain postive HER2 immunostain
negative and/or SISH negative breast cancers are included. Data of
invasive ductal carcinomas were used for training set and data of
invasive lobular carcinoma were used for test set. Number of
observations were 530 in training set and 188 in test set.

@@

To narrow down potential predictors, Genes with a large the median
absolute deviation (more than third-quartiles) were selected. 5000 out
of 20502 genes were included in the modeling process. Yeo-Johnson
transformation was done to correct skewness. Centering and scaling were
done. All preprocessing was done using recipe r package. {[}5{]}
Penalized logistic regression was applied to prediction modeling.
10-fold cross-validation with targe variable stratification was done
over the hyperparameter grid: \(\lambda\) \{\(10^{-5}\),
\(10^{-4}\),\(10^{-3}\),\(10^{-2}\),\(10^{-1}\), \(10^{0}\)\},
\(\alpha\) \{0.0, 0.25, 0.5, 0.75\}. Model performance was evealuated
with the area under receiver operating characteristic (ROC) and area
under the ROC (AUROC) and precision recall (PR) and area under curve
(AUPR). The modelin process was done with tidymodels rpackage. {[}6{]}

@@

\hypertarget{selecting-model}{%
\subsection{Selecting model}\label{selecting-model}}

Cross-validation\\
The model showed best performance at lambda = 0.01 and alpha = 1.0
(Ridge regression).

\hypertarget{prediction-performance}{%
\subsection{Prediction performance}\label{prediction-performance}}

@@

PIK3CA prediction model performed better than RAS activation prediction
model in terms of AUROC. AUPR and the PIK3CA mutation rate
interpretation. {[}7{]}

\begin{itemize}
\tightlist
\item
  Conclusions comments on possible biological or clinical implications
  and suggestions for further research.
\end{itemize}

>>>>

We used TCGA pancancer dataset. PIK3CA mutation data was get using cgdsr
rpackage.{[}3{]} Gene expression data was get from GDAC firehose using
RTCGAToolbox R package. {[}4{]}

@@

To narrow down potential predictors, Genes with a large the median
absolute deviation (more than third-quartiles) were selected. 5000 out
of 20502 genes were included in the modeling process. Yeo-Johnson
transformation was done to correct skewness. Centering and scaling were
done. All preprocessing was done using recipe r package. {[}5{]}
Penalized logistic regression was applied to prediction modeling.
10-fold cross-validation with targe variable stratification was done
over the hyperparameter grid: \(\lambda\) \{\(10^{-5}\),
\(10^{-4}\),\(10^{-3}\),\(10^{-2}\),\(10^{-1}\), \(10^{0}\)\},
\(\alpha\) \{0.0, 0.25, 0.5, 0.75\}. Lambda is penalty scaling parameter
and alpha is mixing parameter of penalty function
(\((1-\alpha)/2 \lVert\beta\rVert_2^2+\alpha\lVert \beta \rVert_1\)).
Model performance was evealuated with the area under receiver operating
characteristic (ROC) and area under the ROC (AUROC) and precision recall
(PR) and area under curve (AUPR). The modelin process was done with
tidymodels rpackage. {[}6{]}

@@

\hypertarget{selecting-model-and-performance-estimation}{%
\subsection{Selecting model and performance
estimation}\label{selecting-model-and-performance-estimation}}

In 10-fold cross-validation, the model with lambda = 0.01 and alpha =
1.0 (Ridge regression) were best performance in terms of AUROC. The
final model was trained with the selected hyperparameters with all
trainset.

@@

Our PIK3CA prediction model performed better than RAS activation
prediction model of previous study in terms of both AUROC (0.84 vs 0.75)
and AUPR (0.39 vs 0.24) on testset prediction. Our testset is
corresponding to the samples initially filtered from training. The
target variable of our study is more specific than the previous study.
The specific important mutations can effect stronger downsteam gene
expression than the broad events pathway activation. The previous study
might be more difficult prediction problum.

OUr model includes cancer type predictor and they are stronger than gene
expression data. The varing prevalence of PIC3CA mutation across cancer
type might reason of the strong cancer type predictor. If the cancer
type was wrong or can not be determined, our model performance can be
poor.

\begin{itemize}
\item
  Conclusions comments on possible biological or clinical implications
  and suggestions for further research.
\item
  Proof of concept study\\
\item
  Reproducibility of gene expression prediction model
\end{itemize}
